---
title: "Bayesian dementia"
authors: Faith & Elora
format: html
editor: visual
---

# Determining the strongest predictor variables and interactions for early onset dementia.

#### This is the r markdown for importing our libraries, data set, cleaning and Bayesian modeling. The document is organized with 4 sections, data cleaning, visual EDA, Bayesian modeling/testing, and then comparison and conclusion.

## 1. Libraries, dataset, and data cleaning.

#### A. libraries

```{r}
library(tidyverse)
library(bayesrules)
library(rstanarm)
library(bayesplot)
library(readr)
```

#### B. Importing datasets and Defining variables

Source of datasets: <https://sites.wustl.edu/oasisbrains/home/oasis-1/>

```{r}
data <- read_csv("merged_oasis_data.csv")
data
```

MMSE (Mini-Mental State Examination):

CDR (Clinical

#### C. Data Cleaning

```{r}
oasis <- data %>% 
select(Age, CDR, nWBV, 
         MMSE) %>% 
  na.omit()
```

## 2. Visual EDA/ exploring

## 3. Bayesian modeling

## 4. Comparison and conclusion
