---
title: "Bayesian dementia"
authors: Faith & Elora
format: html
editor: visual
---

# Determining the strongest predictor variables and interactions for early onset dementia.

#### This is the r markdown for importing our libraries, data set, cleaning and Bayesian modeling. The document is organized with 4 sections, data cleaning, visual EDA, Bayesian modeling/testing, and then comparison and conclusion.

## 1. Libraries, dataset, and data cleaning.

#### A. libraries

```{r}
library(tidyverse)
library(bayesrules)
library(rstanarm)
library(bayesplot)
library(readr)
```

#### B. Importing datasets and Defining variables

Source of datasets: <https://sites.wustl.edu/oasisbrains/home/oasis-1/>

```{r}
data <- read_csv("merged_oasis_data.csv")
data
```

The variables we will be using are:

-   Age: The age of different subjects with and without signs of dementia.

-   MMSE (Mini-Mental State Examination): A measure used to account for variations in head size by normalizing brain volume using total intracranial volume (TIV).

-   CDR (Clinical Dementia Rating): The CDR scale is used to stage the severity of dementia, with specific scores indicating different levels of cognitive and functional impairment. 0 = nondemented; 0.5 – very mild dementia; 1 = mild dementia; 2 = moderate dementia.  

-   nWBV (Normalized Whole Brain Volume): A 30-point questionnaire used as a screening tool to assess cognitive impairment, measuring aspects like memory, orientation, attention, and language.

#### C. Data Cleaning

```{r}
oasis <- data %>% 
select(Age, CDR, nWBV, 
         MMSE) %>% 
  na.omit()
```

```{r}
glimpse(oasis)
```

## 2. Visual EDA/ exploring

#### These will just be some simple plots and charts to understand our variables of interest a little more before we get to modeling. It can also serve to help us make inferences before we get started.

#### A. First I have to change the CDR values from numeric to categorical because theres only 4 different values 235 rows. Ill be able to visualize my data a lot better this way.

```{r}
oasis$CDR <- factor(oasis$CDR,
                       levels= c(0, 0.5, 1, 2),
                       labels= c("nondemented","slight dementia", "mild dementia", "moderate dementia"))
```

#### B. Now I can make some visualizations

```{r}
ggplot( data=oasis, aes(x=CDR, y=Age)) +
  geom_violin()
```

```{r}
ggplot( data=oasis, aes(x=Age, y=CDR,  fill=nWBV)) +
  geom_violin(alpha=0.4)+ geom_boxplot(width=0.2)
```

## 3. Bayesian modeling

```{r}

```

## 4. Comparison and conclusion
